### 2013-08-22 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
moving parallel metrics to an appendix
moving mcsa relaxation parameters to an appendix

%%---------------------------------------------------------------------------%%
\section{Derivation of the Monoenergetic $SP_N$ Equations}
\label{sec:spn_equations}
The $P_N$ equations as derived in \cite{lewis_computational_1993}
give $N+1$ coupled first-order equations capturing the spatial and
angular-dependence of the solution:
\begin{equation}
   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
  \label{eq:final_pn_equations}
\end{equation}
where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
yields a set of $N+1$ equations for $N+2$ flux moments. In multiple
dimensions, the equation set becomes large and coupled not only
through angular moments but also through the spatial variables. As a
simpler alternative to multidimensional $P_N$ solutions, Gelbard
recognized in 1960 that the planar $P_N$ equations could be simplified
and applied an ad-hoc method to extend them to multiple dimensions,
yielding the $SP_N$ equations. These equations are not only fewer in
number, but also take on a diffusion-like form while maintaining the
angular character of the flux, making them amenable to solutions with
modern diffusion methods.

@@

Immediately, we note the diffusion-like nature of
Eq~(\ref{eq:reduced_pn}) as compared to the original $P_N$
equations. To extend these equations to multiple dimensions, Gelbard
simply replaced the planar spatial derivatives in the reduced set of

@@

\subsection{Eigenvalue Form of the $SP_N$ Equations}
\label{subsec:eigenvalue_form}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
source in Eq~(\ref{eq:spn_matrix}) with a fission source this time
considering the multigroup form of the $SP_N$ equations derived in
Appendix~\ref{chap:mg_spn_equations}:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
fission matrix, $\mathbf{F}$, given by
\cite{evans_simplified_2013}. We again apply the change of variables
to yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
the application of the fission matrix to the moment vectors will be
expanded into a set of block matrices in identical fashion to the
scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
It should be noted here that with respect to a general MCSA scheme,
the introduction of fission in the system will only affect the source
vector in the linear system for the eigenvalue schemes used in this
work. The linear operator, and therefore overall MCSA performance will
be dictated by streaming and scattering as defined on the left-hand
side.

%%---------------------------------------------------------------------------%%
\section{Basic Preconditioning for the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Previous work in applying MCSA to the thermal radiation diffusion
equations had success in using Jacobi preconditioning
\cite{evans_monte_2012}. In this section, we will compute the spectral
radius of the iteration matrix formed by the $SP_N$ equations while
parametrically varying the $SP_N$ order, the $P_N$ order, the number
of energy groups, and upscattering/downscattering between energy
groups for Jacobi preconditioned problems to ensure that this type of
preconditioning will be applicable to the fuel assembly problem. For
each problem in the parameter study, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.

@@

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively. For these figures, the parameters in
Table~\ref{tab:spn_fixed_parameters} with the number of spatial
elements reduced to $2 \times 2 \times 2$ to show the blocks generated
by the discretization.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.25in]{chapters/spn_equations/group1.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
      used to show detail of the blocks formed by the discretization.}}
  \label{fig:group1}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.25in]{chapters/spn_equations/group10ds.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter only.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10ds}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.25in]{chapters/spn_equations/group10us.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10us}
\end{figure}
We note a few key features of the sparsity plots. The first is that
for multigroup problems without full upscatter and downscatter
(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
and therefore a linear solver that can handle asymmetric linear
systems is required. Nearly all problems of interest will not have
full upscattering or downscattering. Second we note the largely
diagonal character of these systems, although the blocks from
Eq~(\ref{eq:A_block_matrix}) are readily apparent.

Based on this both block and diagonally dominant structure for
matrices formed by the general multigroup $SP_N$ equations, we choose
\textit{block Jacobi} preconditioning as a left preconditioner for the
system. Block Jacobi preconditioning extracts the diagonal elements of
the matrix as the preconditioner where now the elements extracted are
the blocks on the diagonal as shown on the left side of
Figure~\ref{fig:block_jacobi_ex}. Shown on the right side of
Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner is
trivial with each diagonal block inverted separately. For the $SP_N$
equations, Eq~(\ref{eq:A_block_matrix}) gives a block size of
$N_g\times(N+1)/2$.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
  \end{center}
  \caption{\textbf{Block Jacobi preconditioning strategy used for the
      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
      the diagonal blocks of the matrix. Right: Inversion of the
      preconditioner is trivial and decoupled by block.}}
  \label{fig:block_jacobi_ex}
\end{figure}

For high performance implementations this has several attractive
properties. First, the blocks in the matrix come from the energy/angle
discretization of the transport equation as given by
Eq~(\ref{eq:A_block_matrix}). Each block on the diagonal is bound to a
mesh element in the system (note there are 8 blocks on the diagonal in
each of the sparsity patterns with a mesh of $2 \times 2 \times 2$)
and therefore we expect the matrix elements forming the block to be
entirely local in domain decomposed calculations. Second, these blocks
are typically dense and nearly lower triangular for many transport
problems meaning that established dense matrix methods can be used for
fast inversion.

@@

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin, control rod, and moderator materials and
the associated grid is given by Figure~\ref{fig:problem3_radial_mat}
while a cross section of the materials configuration along the
horizontal axis is given by Figure~\ref{fig:problem3_axial_mat}. A
detailed view of the assembly bottom is given in
Figure~\ref{fig:problem3_end}. On the top and bottom of the assembly,
vacuum conditions are used as well as on the top and right boundaries
in Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are
used on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly. For the spatial
discretization, each fuel pin (gray regions in
Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
mesh with materials and cross sections homogenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized control rods. The control rods were removed for
      these problems. Each fuel pin is resolved by a $2 \times 2$
      mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is moderator and darker
      purple/red material is fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is moderator and darker purple/red material is fuel.}}
  \label{fig:problem3_end}
\end{figure}

@@

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. This
problem is significantly more complicated than the simple test problem
used for the previous spectral analysis. Fission has been introduced
into the set of equations and the addition of moderator into the
system will increase the amount of scattering, creating a
significantly more difficult problem manifesting itself in an
iteration matrix with a larger spectral radius. When using MCSA, the
linear operator applied to $\mathbf{\Phi}^{n+1}$ at each eigenvalue
iteration will dictate convergence and remain unchanged throughout the
eigenvalue computation while the addition of fission to the system
will only modify the source of neutrons and the multiplication factor
at every eigenvalue iteration while not affecting Monte Carlo
transport.

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
We first use block Jacobi preconditioning to solve the $17 \times 17$
fuel assembly problem. A single energy group problem was first
computed with $SP_1$ discretization, effectively giving the one-speed
neutron diffusion system for the fuel assembly resulting in 20,088
degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. Even if
the problem may be ill-conditioned, we do expect convergence of MCSA
given that the spectral radius requirement was met. To investigate
further, a block Jacobi preconditioned Richardson iteration was used
to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor convergence is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

Based on these results, it then appears that even if the simple
criteria of a spectral radius of less than one is met and the
Richardson iteration will converge with the same preconditioning, MCSA
still may not converge. We then expect the issue to reside with the
Neumann-Ulam solve providing the correction as the Richardson
iteration is known to provide the correct result. Furthermore, the
fact that the spectral radius is less than one means that the
stochastic histories in the block Jacobi preconditioned Neumann-Ulam
method are eventually being terminated by the weight cutoff as no
artificial absorption was used for these preliminary
calculations. Based on this, the correction being generated by the
Neumann-Ulam solve is the component of MCSA causing a divergent
solution.

@@

%%---------------------------------------------------------------------------%%
\section{Advanced Preconditioning Strategies}
\label{subsec:spn_advanced_preconditioning}
For the fuel assembly criticality problem presented in the previous
section, the spectral radius was near one using Jacobi preconditioning
and in a regime in which MCSA breakdown is observed. In this regime
the number of stochastic histories required to converge MCSA increases
rapidly and the resulting poor performance is compounded by those
histories being increasingly expensive to compute. To overcome this
difficulty, advanced preconditioning strategies for the $SP_N$ problem
are required beyond simple Jacobi methods that can reduce the spectral
radius into a region of better MCSA performance. Two modern algebraic
preconditioning strategies, ILUT and SPAINV, will be presented here
and used within the explicit preconditioning framework given by
Eq~(\ref{eq:left_right_mcsa}). Data
showing their effects on MCSA solutions of the fuel assembly
criticality problem will be presented.

>>>>

%%---------------------------------------------------------------------------%%
\section{Derivation of the Monoenergetic $SP_N$ Equations}
\label{sec:spn_equations}
The $P_N$ equations as derived in \cite{lewis_computational_1993}
give $N+1$ coupled first-order equations capturing the spatial and
angular-dependence of the solution:
\begin{equation}
   \frac{1}{2n+1} \frac{\partial}{\partial x}\Big[ (n+1) \phi_{n+1} + n
     \phi_{n-1} \Big] + \Sigma_n \phi_n = q\delta_{n0} \:,
  \label{eq:final_pn_equations}
\end{equation}
where $\Sigma_n = \sigma-\sigma_{sn}$ and the summations are truncated
at some level of approximation $N$ such that $n = 0,1,\dotsc,N$. This
yields a set of $N+1$ equations for $N+2$ flux moments with the
highest order moment set to zero for closure. In multiple dimensions,
the equation set becomes large and coupled not only through angular
moments but also through the spatial variables. As a simpler
alternative to multidimensional $P_N$ solutions, Gelbard recognized in
1960 that the planar $P_N$ equations could be simplified and applied
an ad-hoc method to extend them to multiple dimensions, yielding the
$SP_N$ equations. These equations are not only fewer in number, but
also take on a diffusion-like form while maintaining the angular
character of the flux, making them amenable to solutions with modern
diffusion methods.

@@

To extend these equations to multiple dimensions, Gelbard simply
replaced the planar spatial derivatives in the reduced set of

@@

\subsection{Eigenvalue Form of the $SP_N$ Equations}
\label{subsec:eigenvalue_form}
For criticality problems, the space, angle, and energy discretizations
for the $SP_N$ equations can be applied to the general eigenvalue form
of the transport equation given by
Eq~(\ref{eq:eigenvalue_transport}). We can readily replace the fixed
source in Eq~(\ref{eq:spn_matrix}) with a fission source this time
considering the multigroup form of the $SP_N$ equations derived in
Appendix~\ref{chap:mg_spn_equations}:
\begin{multline}
  -\nabla \cdot \Bigg[\frac{n}{2n+1}\mathbf{\Sigma_{n-1}}^{-1} \nabla
    \Big(\frac{n-1}{2n-1} \mathbf{\Phi_{n-2}} +
    \frac{n}{2n-1}\mathbf{\Phi_n} \Big) \\+
    \frac{n+1}{2n+1}\mathbf{\Sigma_{n+1}}^{-1} \nabla
    \Big(\frac{n+1}{2n+3}\mathbf{\Phi_n} +
    \frac{n+2}{2n+3}\mathbf{\Phi_{n+2}}\Big) \Bigg] \\+
  \mathbf{\Sigma_n} \mathbf{\Phi_n} = \frac{1}{k} \mathbf{F}
  \mathbf{\Phi_n} \delta_{n0} \ \ \ \ \ \ \ \ \ n = 0,2,4,\cdots,N\:.
  \label{eq:multigroup_spn_eigenvalue}
\end{multline}
with $\mathbf{\Phi_n}$ the vector of multigroup flux moments and the
fission matrix, $\mathbf{F}$, given by
\cite{evans_simplified_2013}. We again apply the change of variables
to yield the set of multigroup pseudo-moments $\mathbb{U}_n$ where now
the application of the fission matrix to the moment vectors will be
expanded into a set of block matrices in identical fashion to the
scattering matrices:
\begin{equation}
  -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
  \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
  \mathbb{F}_{nm} \mathbb{U}_m\ \ \ \ \ \ \ n = 1,2,3,4\:.
  \label{eq:spn_fission_matrix}
\end{equation}
To complete the spatial discretization, the gradient operators are
resolved by a finite volume scheme as in
\cite{evans_simplified_2013}. It should be noted here that with
respect to a general MCSA scheme, the introduction of fission in the
system will only affect the source vector in the linear system for the
eigenvalue schemes used in this work. The linear operator, and
therefore overall MCSA performance will be dictated by streaming and
scattering as defined on the left-hand side.

%%---------------------------------------------------------------------------%%
\section{Basic Preconditioning for the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Previous work in applying MCSA to the thermal radiation diffusion
equation had success in using Jacobi preconditioning
\cite{evans_monte_2012} and therefore we apply it here to the $SP_N$
equations given their diffusion-like form. In this section, we will
compute the spectral radius of the iteration matrix formed by the
$SP_N$ equations while parametrically varying the $SP_N$ order, the
$P_N$ order, the number of energy groups, and
upscattering/downscattering between energy groups for Jacobi
preconditioned problems to ensure that this type of preconditioning
will be applicable to the fuel assembly problem. For each problem in
the parameter study, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.

@@

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively. For these figures, the parameters in
Table~\ref{tab:spn_fixed_parameters} with the number of spatial
elements reduced to $2 \times 2 \times 2$ to show the blocks generated
by the discretization.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.3in]{chapters/spn_equations/group1.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
      used to show detail of the blocks formed by the discretization.}}
  \label{fig:group1}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.3in]{chapters/spn_equations/group10ds.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter only.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10ds}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3.3in]{chapters/spn_equations/group10us.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10us}
\end{figure}

We note a few key features of the sparsity plots. The first is that
for multigroup problems without full upscatter and downscatter
(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
and therefore a linear solver that can handle asymmetric linear
systems is required. Nearly all problems of interest will not have
full upscattering or downscattering. Second we note the largely
diagonal character of these systems, although the blocks from
Eq~(\ref{eq:A_block_matrix}) are readily apparent with a block size of
$N_g\times(N+1)/2$.

Based on this both block and diagonally dominant structure for
matrices formed by the general multigroup $SP_N$ equations, we choose
\textit{block Jacobi} preconditioning as a left preconditioner for the
system. Block Jacobi preconditioning extracts the diagonal block
elements of the matrix to use as the preconditioner as shown on the
left side of Figure~\ref{fig:block_jacobi_ex}. Shown on the right side
of Figure~\ref{fig:block_jacobi_ex}, inversion of the preconditioner
is trivial with each diagonal block inverted separately.

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.6}{
    \input{chapters/spn_equations/block_jacobi.pdftex_t} }
  \end{center}
  \caption{\textbf{Block Jacobi preconditioning strategy used for the
      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
      the diagonal blocks of the matrix. Right: Inversion of the
      preconditioner is trivial and decoupled by block.}}
  \label{fig:block_jacobi_ex}
\end{figure}

For high performance implementations block Jacobi preconditioning has
several attractive properties. First, the blocks in the transport
operator come from the energy/angle discretization of the transport
equation as given by Eq~(\ref{eq:A_block_matrix}). Each block on the
diagonal is bound to a mesh element in the system (note there are 8
blocks on the diagonal in each of the sparsity patterns with a mesh of
$2 \times 2 \times 2$) and therefore we expect the matrix elements
forming the block to be entirely local in domain decomposed
calculations. Second, these blocks are typically dense and nearly
lower triangular for many transport problems meaning that established
dense matrix methods can be used for fast inversion.

@@

%%---------------------------------------------------------------------------%%
\section{Fuel Assembly Criticality Calculations}
\label{sec:fuel_assembly_calcs}
Fuel assembly calculations are a critical piece of nuclear engineering
infrastructure for reactor core analysis and design. At this level,
individual fuel pins may be resolved at fine resolution in a variety
of configurations. As a sophisticated problem of interest to push the
limits of MCSA, a hot zero-power $17 \times 17$ pin assembly will be
used with varying energy group structure and $SP_N$ discretization in
a criticality calculation. A cross section along the vertical axis
showing homogenized fuel pin, control rod, and moderator materials and
the associated grid is given by Figure~\ref{fig:problem3_radial_mat}
while a cross section of the materials configuration along the
horizontal axis is given by Figure~\ref{fig:problem3_axial_mat}. The
control rods were removed for these problems leaving only moderator in
their place. A detailed view of the assembly bottom is given in
Figure~\ref{fig:problem3_end}. On the top and bottom of the assembly,
vacuum conditions are used as well as on the top and right boundaries
in Figure~\ref{fig:problem3_radial_mat}. Reflecting conditions are
used on the left and bottom boundaries of
Figure~\ref{fig:problem3_radial_mat}, effectively giving a
representation of one quarter of the assembly. For the spatial
discretization, each fuel pin (gray regions in
Figure~\ref{fig:problem3_radial_mat}) is resolved by a $2 \times 2$
mesh with materials and cross sections homogenized over this region.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/problem3_radial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly mesh and geometry cross section.}
    \textit{Reflecting boundaries are used on the left and lower
      boundaries to create a complete $17 \times 17$ assembly
      geometry. Gray regions are homogenized fuel and red regions are
      homogenized control rods. Each fuel pin is resolved by a $2
      \times 2$ mesh.}}
  \label{fig:problem3_radial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_axial_mat.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry cross section.} \textit{The
      geometry is subdivided along the axial direction into 50 zones
      spaced to capture material boundaries. Important details include
      spacer grids along the length of the fuel pins and reactor core
      structures on the top and bottom of the assembly. Lighter purple
      material in the center of the assembly is homogenized
      moderator/control rodes and darker purple/red material is
      homogenized moderator/fuel.}}
  \label{fig:problem3_axial_mat}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6.0in]{chapters/spn_equations/problem3_end.png}
  \end{center}
  \caption{\textbf{Fuel assembly geometry end detail.}
    \textit{Reactor core structure including spacer grids and plenum
      has been included. Lighter purple material on the right of the
      figure is homogenized moderator/control rods and darker
      purple/red material is homogenized moderator/fuel.}}
  \label{fig:problem3_end}
\end{figure}

@@

To generate the multiplication factor and steady-state flux
distribution for this problem, at every eigenvalue iteration MCSA is
used to solve the resulting $SP_N$ problem using the provided fission
source. Algorithm~\ref{alg:power_iteration} presents the use of MCSA
within a power iteration strategy to find the multiplication factor.
\begin{algorithm}[h!]
  \caption{Power Iteration MCSA Scheme}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State $k_0 =$ initial guess
    \State $\mathbf{\Phi}_0 =$ initial guess
    \State $n = 0$
    \While{$|\frac{k^n - k^{n-1}}{k^n}| < \epsilon$}
    \Comment{Iterate until convergence of the eigenvalue}
    \State $\mathbf{M} \mathbf{\Phi}^{n+1} = \frac{1}{k^n} \mathbf{F} \mathbf{\Phi}^n$
    \Comment{Solve for the new flux state with MCSA}
    \State $k^{n+1} = k^n \frac{\int \mathbf{F} \mathbf{\Phi}^{n+1} d\mathbf{r}}{\int
      \mathbf{F} \mathbf{\Phi}^n d\mathbf{r}}$
    \Comment{Update the multiplication factor}
    \State $n = n+1$
    \EndWhile
  \end{algorithmic}
\end{algorithm}
Here, $\mathbf{M}$ is the transport operator generated on the
left-hand side of the $SP_N$ discretization, $\mathbf{F}$ is the
fission matrix, and $\mathbf{\Phi}$ the multigroup neutron flux. Using
MCSA in this context is significantly more complicated than the simple
test problem used for the previous fixed-source preconditioner
analysis. Fission has been introduced into the set of equations and
the addition of moderator into the system will increase the amount of
scattering, creating a significantly more difficult problem
manifesting itself in an iteration matrix with a larger spectral
radius. When using MCSA, the linear operator applied to
$\mathbf{\Phi}^{n+1}$ at each eigenvalue iteration will dictate
convergence and remain unchanged throughout the eigenvalue computation
while the addition of fission to the system will only modify the
source of neutrons and the multiplication factor at every eigenvalue
iteration while not affecting Monte Carlo transport.

\subsection{Preliminary Jacobi Preconditioned Calculations}
\label{subsec:jacobi_prec_assembly_calc}
We first use block Jacobi preconditioning to solve the $17 \times 17$
fuel assembly problem. A single energy group problem was first
computed with $SP_1$ discretization, effectively giving the one-speed
neutron diffusion system for the fuel assembly resulting in 20,088
degrees of freedom in the
problem. Figure~\ref{fig:block_jacobi_res_mcsa} gives the residual
infinity norm as a function of iteration for the MCSA linear solve in
the first eigenvalue iteration using 25,000 stochastic histories at
every iteration for the adjoint Neumann-Ulam solve.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned MCSA solve during the first eigenvalue
      iteration of the 1-group $17 \times 17$ fuel assembly problem.}
    \textit{Convergence was not achieved with the block Jacobi
      preconditioned method.}}
  \label{fig:block_jacobi_res_mcsa}
\end{figure}
Convergence was not achieved as noted by the rapid rise in the
residual over a few iterations. Based on the spectral radius
computations performed, these results are not in line with
expectations for this problem. Additional computations performed with
\sn{1}{6} histories per iteration exhibited the same divergent
behavior at a significant computational cost and compute time. To
investigate further, a block Jacobi preconditioned Richardson
iteration was used to solve the same
problem. Figure~\ref{fig:block_jacobi_res_richardson} gives the
residual infinity norm as a function of iteration for the Richardson
linear solve in the first eigenvalue iteration.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/block_jacobi_rich_res.pdf}
  \end{center}
  \caption{\textbf{Residual infinity norm vs. iteration for the block
      Jacobi preconditioned Richardson solve during the first
      eigenvalue iteration of the 1-group $17 \times 17$ fuel assembly
      problem.} \textit{A spectral radius near 1 was observed for the
      iteration matrix.}}
  \label{fig:block_jacobi_res_richardson}
\end{figure}
Poor convergence is observed for the Richardson iteration, however,
convergence is achieved meaning that the preliminary eigenvalue
criteria needed to satisfy MCSA convergence has also been met. The
number of iterations required for the Richardson iteration to converge
will give an approximation for the spectral radius via
Eq~(\ref{eq:linear_k_iter_norm3}). With 7291 iterations required to
converge to a tolerance of $\sn{1}{-6}$, $\rho(\mathbf{H}) \approx
0.998$, nearing the limits of MCSA applicability and well beyond the
spectral radii generated in the initial spectral analysis.

Based on these results, it then appears that even if the simple
criteria of a spectral radius of less than one is met and the
Richardson iteration will converge with the same preconditioning, MCSA
still may not converge. We then expect the issue to reside with the
Neumann-Ulam solve providing the correction as the Richardson
iteration is known to provide the correct result. Furthermore, the
fact that the spectral radius is less than one means that the
stochastic histories in the block Jacobi preconditioned Neumann-Ulam
method are eventually being terminated by the weight cutoff as no
artificial absorption was used for these preliminary
calculations. Based on this, the correction being generated by the
Neumann-Ulam solve is the component of MCSA causing a divergent
solution which we study next.

@@

%%---------------------------------------------------------------------------%%
\section{Advanced Preconditioning Strategies}
\label{subsec:spn_advanced_preconditioning}
For the fuel assembly criticality problem presented in the previous
section, the spectral radius was near one using Jacobi preconditioning
and in a regime in which MCSA breakdown is observed. In this regime
the number of stochastic histories required to converge MCSA increases
rapidly and the resulting poor performance is compounded by those
histories being increasingly expensive to compute. To overcome this
difficulty, advanced preconditioning strategies for the $SP_N$ problem
are required beyond simple Jacobi methods that can reduce the spectral
radius into a region of better MCSA performance. Two modern algebraic
preconditioning strategies, ILUT and SPAINV, will be presented here
and used within the explicit preconditioning framework given by
Eq~(\ref{eq:left_right_mcsa}). Data showing their effects on MCSA
solutions of the fuel assembly criticality problem will be provided.
